Seasonality in governing upper-ocean dynamics at submesoscales

### 2016-05-09 writeup/rocha_etal.tex
Fixes typos

% Shorter version of title entered in capital letters:
\titlerunninghead{Seasonality in dominant upper-ocean dynamics at submesoscales}

@@

\title{Seasonality in dominant upper-ocean dynamics at submesoscales}
%
% e.g., \title{Terrestrial ring current:
% Origin, formation, and decay $\alpha\beta\Gamma\Delta$}
%

@@

\section{The LLC2160 model output}
We use the output of the latitude-longitude polar cap (LLC)
MITgcm  numerical simulation. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and high-resolution surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years form March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.

@@

Focusing on the lateral gradients of the horizontal velocity vector.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$ and
horizontal divergence $\delta \equiv u_x + v_y$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

@@

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: low-pass filtering the velocity field
with a cut-off scale of 100 km reduces the RMS vorticity by at least 40$\%$,
yielding bulk statistics of velocity gradient roughly consistent with AVISO
geostrophic velocity data (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows project are supra-inertial and project
on scales smaller than 100 km: daily-averaging the velocity fields or low-pass
filtering (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with that
of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

To better understand the dynamics the projection of these flows on horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and any spatial linear trend were
removed before calculating discrete Fourier transforms.Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales small than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100km, 40$\%$ of the surface KE in April a is accounted for by suprainertial
flows as opposed to 77$\%$ in October. The seasonal of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield no significant seasonality in the spectra
of hourly SSH.

@@

\section{Projection on time scales}
To better understand the high-frequency/high-wavenumber flows  we calculate frequency
spectra of surface KE (figure \ref{fig4}). To obtain statistically meaningful results,
we average spectral estimates along a section at 165$^\circ$E.  There is also about 50
independent realizations of spectral estimates.

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominates by inertia-gravity
waves. (Note that the the inertial peak nearly merges with the
diurnal tide peak). There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former
are more energetic in April and the latter in October.

Most of these high-frequency inertia-gravity waves project onto scales smaller
than 100 km. The surface KE spectra of 100-km-smoothed velocity field is significantly
suppressed in supra-inertial variability --- there are no statistically significant
changes at sub-inertial frequencies (see dashed lines in figure \ref{fig4}).

@@

\section{Summary and Discussion}
Our main finding is that
upper-ocean divergent flows, dominated by inertia-gravity waves, undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the quasi-balanced flows. Therefore, the present results suggest significant
seasonal modulation of the dominant upper-ocean dynamics at submesoscales
(10-100 km): Quasi-balanced turbulence driven by shallow baroclinic instabilities
 in late Winter/early Spring
and inertia-gravity waves dominate in late Summer/early Fall.

That the surface velocity and SSH submesoscale (10-100 km) variability may be
dominated by ageostrophic flows in
Summer/Fall has consequences for the interpretation of data from
the future SWOT and COMPIRA altimeter missions,
which will deliver SSH measurements at submesoscales. To the extent that
high-frequency flows are dominated by incoherent internal tides and other
internal waves, it may be very difficult (if not impossible) to
separate submesoscale SSH variability associated with geostrophic motions
from high-frequency, ageostrophic flows. Thus, previous claims that
one will be able to easily obtain submesoscale surface
geostrophic velocities and monitor such seasonal cycle on global scales
\citep{sasaki_etal2014,qiu_etal2014} are overstated.

The present analysis focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results with care.  The effects of smaller-scale/higher-frequency
``sub-submesoscale'' flows on the submesoscale surface velocity and SSH variability
are presently unknown.

>>>>

% Shorter version of title entered in capital letters:
\titlerunninghead{Seasonality in governing upper-ocean dynamics at submesoscales}

@@

\title{Seasonality in governing upper-ocean dynamics at submesoscales}
%
% e.g., \title{Terrestrial ring current:
% Origin, formation, and decay $\alpha\beta\Gamma\Delta$}
%

@@

\section{The LLC2160 model output}
We use the output of a latitude-longitude polar cap (LLC)
MITgcm  numerical simulation  \citep{forget_etal2015}. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and 6-hourly surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years form March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution  of the
LLC2160 simulation is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.
 (see figure \ref{fig3}).

@@

Focusing on the submesoscale variability, we present estimates of bulk quantities
associated with the the horizontal velocity lateral gradient.
The lateral gradients are calculated using a centered
second-order finite differences scheme. We then calculate
the vertical vorticity $\zeta \equiv v_x - u_y$ and
horizontal divergence $\delta \equiv u_x + v_y$.
These quantities highlight the submesoscale structures in the flow
and hint on their dynamics.

@@

The bulk of the vertical vorticity is associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields reduces the vertical
vorticity by less than 20$\%$ and the seasonal cycle remains strong (see
green lines in figure \ref{fig2}). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity field
with a cut-off scale of 100 km reduces the RMS vorticity by at least 40$\%$,
yielding bulk statistics of velocity gradient roughly consistent with AVISO
geostrophic velocities (see lines blue and black in figure \ref{fig2}).

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows are supra-inertial and project
on scales smaller than 100 km: daily-averaging or smoothing the velocity fields
 (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with the
seasonal cycle of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
\vspace{-1.5cm}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

To better understand the projection of these flows onto different horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and spatial linear trend were
removed before calculating discrete Fourier transforms.Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales small than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100 km, 40$\%$ of the surface KE in April a is accounted for by suprainertial
flows as opposed to 77$\%$ in October. The seasonality of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100 km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield no significant seasonality in the spectra
of hourly SSH.

@@

\section{Projection on time scales}
To better understand the high-frequency/high-wavenumber flows  we calculate frequency
spectra of surface KE (figure \ref{fig4}). To obtain statistically meaningful results,
we average spectral estimates along a section at 165$^\circ$E.  There is about 50
independent spectral realizations.

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominated by inertia-gravity
waves. (Note that the inertial peak nearly merges with the
diurnal tide peak.) There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former
are more energetic in April and the latter in October.

Most of these high-frequency inertia-gravity waves project onto scales smaller
than 100 km. The surface KE spectra of 100-km-smoothed velocity field is significantly
suppressed in supra-inertial frequencies --- there are no statistically significant
changes at sub-inertial frequencies (see dashed lines in figure \ref{fig4}).

@@

\section{Summary and Discussion}
Our main finding is that
upper-ocean divergent flows, dominated by inertia-gravity waves, undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
the quasi-balanced flows. Therefore, the present results suggest significant
seasonal modulation of the dominant upper-ocean dynamics at submesoscales
(10-100 km): quasi-balanced turbulence driven by shallow baroclinic instabilities
 in late Winter/early Spring
and inertia-gravity waves dominate in late Summer/early Fall.

That the surface velocity and SSH submesoscale (10-100 km) variability may be
dominated by ageostrophic flows in
Summer/Fall has consequences for the interpretation of data from
the future SWOT and COMPIRA altimeter missions,
which will deliver SSH measurements at submesoscales. To the extent that
high-frequency flows are dominated by spatially incoherent internal tides and other
internal waves, it may be very difficult (if not impossible) to
separate submesoscale SSH variability associated with geostrophic motions
from high-frequency, ageostrophic flows. Thus, previous claims that
one will be able to easily obtain submesoscale surface
geostrophic velocities and monitor such seasonal cycle on global scales
\citep{sasaki_etal2014,qiu_etal2014} are overstated.

The present analysis focuses on a single patch of ocean in the
vicinities of the Kuroshio Extension, not
representative of other regions such as eastern boundary currents
and the middle of the subtropical gyre: care must be taken in overgeneralizing
these results.  The effects of smaller-scale/higher-frequency
``sub-submesoscale'' flows on the submesoscale surface velocity and SSH variability
are presently unknown.
